Numerical simulation of bubble deformation \\ and breakup under simple linear shear flows

### 2022-12-24 Bubble-deformation.tex
data ready to be plotted
predicted variance of computed data very small

In contrast, few computational studies on bubble deformation and breakup in
shear flow have been conducted \cite{WanShiZha15}.  This type of bubble
dynamics in shear flow is of critical importance for a variety of
scientific and engineering processes. We refer the reader to the 
following experimental studies relating to bubble deformation in foaming
processes, microfluidic devices, and micro bubbles in the blood 
circulation system,
\cite{CHU2019108,muller2008single,bento2018deformation,drenckhan2015science}.
In particular, it is the study of bubble deformation as it pertains
to high performance plastics applications which motivates this work.  
In this article, we present
computational studies of shear-driven deformation and breakup of a bubble in
insoluble viscous liquids.  The physical properties that distinguish bubble and
drop studies are expressed in terms of the density ratio $\lambda = \rho_b /
\rho_m$ and the viscosity ratio $\eta = \mu_b / \mu_m$, where $\rho$ is the
fluid density, $\mu$ is the viscosity and the subscripts ``b'' and ``m'' denote
the ``bubble'' or ``drop'' and the ``matrix fluid'', respectively.  For a
bubble in an insoluble viscous liquid, the density ratio $\lambda \simeq 0$ and
the viscosity ratio $\eta \simeq 0$, whereas most studies dealing with a drop
in an immiscible viscous liquid take $\lambda =1$ and $\eta \simeq 1$.

@@

Table~\ref{tab:CaRecComparison} 
lists, for representative $Ca$ values,
the corresponding critical Reynolds number, $Re_{c}$, for shear-induced 
bubble breakup for a variety of flow
conditions.  The results indicate that sufficiently large shear forces are
required for bubble breakup even for large Capillary numbers. The combined
effects of very small density and viscosity ratios 
($\lambda$ = $\eta$ = 1.0 $\times$ $10^{-3}$) 
act synergistically in deformation and breakup processes.
We conducted a linear least squares best fit analysis (linear vs. 
power law vs. exponential models) of the data that we 
list in Table~\ref{tab:CaRecComparison} and have found that the following
exponential model is the best fit:
%  Re = \frac{\rho_{b}UR}{\mu_{b}}, \quad
%  Ca = \frac{\mu_{b}U}{\sigma}, \quad
\begin{eqnarray}
Re_{c}=4.9 e^{-1.41 (Ca)} \label{bestfit}
\end{eqnarray}
In Figure~\ref{fig:CaRecFit} we plot our best fit curve and make the
hypothesis that a given $(Ca,Re)$ pair will result in bubble break up
if the point $(Ca,Re)$ is above the critical curve.

@@

>>>>

In contrast, few computational studies on bubble deformation and breakup in
shear flow have been conducted \cite{WanShiZha15}.  This type of bubble
dynamics in shear flow is of critical importance for a variety of
scientific and engineering processes. We refer the reader to the 
following experimental studies relating to bubble deformation in foaming
processes, microfluidic devices, and micro bubbles in the blood 
circulation system,
\cite{CHU2019108,muller2008single,bento2018deformation,drenckhan2015science}.
In particular, it is the study of bubble deformation as it pertains
to high performance plastics applications which motivates this work.  
In this article, we present
computational studies of shear-driven deformation and breakup of a bubble in
insoluble viscous liquids.  
Studying bubble break-up via computation rather
than experiments simplifies the process of setting a combination of
precise simple shear flow conditions, low $Ca$ conditions, low density ratio, 
and low viscosity ratios.
The physical properties that distinguish bubble and
drop studies are expressed in terms of the density ratio $\lambda = \rho_b /
\rho_m$ and the viscosity ratio $\eta = \mu_b / \mu_m$, where $\rho$ is the
fluid density, $\mu$ is the viscosity and the subscripts ``b'' and ``m'' denote
the ``bubble'' or ``drop'' and the ``matrix fluid'', respectively.  For a
bubble in an insoluble viscous liquid, the density ratio $\lambda \simeq 0$ and
the viscosity ratio $\eta \simeq 0$, whereas most studies dealing with a drop
in an immiscible viscous liquid take $\lambda =1$ and $\eta \simeq 1$.

@@

Table~\ref{tab:CaRecComparison} 
lists, for representative $Ca$ values,
the corresponding critical Reynolds number, $Re_{c}$, for shear-induced 
bubble breakup for a variety of flow
conditions.  The results indicate that sufficiently large shear forces are
required for bubble breakup even for large Capillary numbers. The combined
effects of very small density and viscosity ratios 
($\lambda$ = $\eta$ = 1.0 $\times$ $10^{-3}$) 
act synergistically in deformation and breakup processes.
We conducted a linear least squares best fit analysis (linear vs. 
power law vs. exponential models) of the data that we 
list in Table~\ref{tab:CaRecComparison} and have found that the following
exponential model is the best fit:
%  Re = \frac{\rho_{b}UR}{\mu_{b}}, \quad
%  Ca = \frac{\mu_{b}U}{\sigma}, \quad
\begin{eqnarray}
Re_{c}=138.3 e^{-1.41 (Ca)} \label{bestfit}
\end{eqnarray}
(\ref{bestfit}) was derived from the linear least squares best fit
of the log of $Re_{c}$:
\begin{eqnarray}
	\log(Re_{c})=4.9-1.41 (Ca)  \label{bestfitraw}
\end{eqnarray}
%% variance for X^T X c = b
%% (S/(n-m)) (X^T X)^{-1}
%% standard deviation: sqrt(2.1E-3)=.05 and sqrt(4.2E-3)=0.06
The predicted variances for the parameters 
``$4.9$'' and ``$-1.41$'' which appear in (\ref{bestfitraw}) are
$2.1E-3$ and $4.2E-3$ respectively.  In other words, even with just
four data points, we have computed a reasonable best fit curve.

In Figure~\ref{fig:CaRecFit} we plot our best fit curve and make the
hypothesis that a given $(Ca,Re)$ pair will result in bubble break up
if the point $(Ca,Re)$ is above the critical curve.

@@

%% variance for X^T X c = b
%% (S/(n-m)) (X^T X)^{-1}
\begin{figure}%[h!]
  \centering
  \includegraphics[width=\textwidth]{critical_Reynolds_Number_label}
  \caption{A plot of the critical Reynolds number versus Capillary number
	for the data listed in Table \ref{tab:CaRecComparison}.  The
	linear least squares best fit for the $\log(Re_{c})$ is
	$\log(Re_{c})=4.9-1.41(Ca)$ (i.e. $Re_{c}=138.3 e^{-1.41 Ca}$)
	with a predicted variance
	for the parameters ``$4.9$'' and ``$-1.41$'' being
	$2.1E-3$ and $4.2E-3$ respectively.  Even for the
	given data that we computed, it is suggested that 
	a ``bubble break-up map''
	can be derived, in which for the conditions that
	$\lambda$ = $\eta$ = 1.0 $\times$ $10^{-3}$ and $0<Ca<1$, one
	can predict whether a given $(Ca,Re)$ pair will result in shear
	induced bubble break-up or not.  
	   }
  \label{fig:CaRecFit}
\end{figure}
%
